noam_data_exploration_v1

### 2024-07-22 noam_data_exploration_v3.Rmd
Continuing to update/play around with models and visualization.

(richness_quad_plot <-
  ggplot(data = alpha_diversity_quad_means) +
  geom_point(aes(x = Year, y = Richness_mean, colour = Island_side, shape = Habitat, group = Island_side_habitat), size = 8) +
  geom_line(aes(x = Year, y = Richness_mean, colour = Island_side, group = Island_side_habitat), size = 1.2) +
  geom_errorbar(aes(x = Year, ymin = Richness_mean - Richness_se , ymax = Richness_mean + Richness_se, colour = Island_side,
                    group = Island_side_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = island_side_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Habitat)+
  labs(title = "a. Richness quad level\n", y = "Richness\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )

@@

slope_mod <- glmmTMB(Slope ~ Habitat*Island_side + (1|Site/Transect), data = gls_coefficients_merged)
summary(slope_mod) # pretty low variance in random effects
plot(gls_res <- simulateResiduals(slope_mod)) # there are issues, but I can't figure out what they are...
# I think this is due to 1200 obs in model
hist(residuals(slope_mod))
plot(predict(slope_mod) ~ residuals(slope_mod))
Anova(slope_mod)
#                        Chisq Df Pr(>Chisq)    
# Island_side           16.075  2  0.0003232 ***
# Habitat             1301.260  3  < 2.2e-16 ***
# Island_side:Habitat  799.995  6  < 2.2e-16 ***
r.squaredGLMM(slope_mod) # 0.16, 0.18
plot(allEffects(slope_mod))
# Backreef: richness increases on East and North shore, but relatively constant west side
# Fringe: richness decreasing on east and west side, but relatvely constant on north shore
# Outer 10: Richness neutral/decreasing on east side, but neutral/increasing on north and west side
# Outer 17: Richness decreasing north/east side, but neutral west side
# Worth noting trends are slight - arguable whether they are biologically meaningful

slope_emm <- emmeans(slope_mod, ~ Habitat*Island_side) # create emmeans object
emmeans::contrast(slope_emm, "consec", simple = "each", combine = TRUE, adjust = "bonferroni")
# Note: not sure I want to include this
```

@@

ggplot(data = gls_coefficients_merged) +
  geom_histogram(aes(x = Slope), bins = 30) +
  geom_vline(xintercept = 0, size = 2, linetype = 2) +
  facet_grid(Habitat ~ Island_side) +
#  theme_classic() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
#        panel.background = element_blank(),
        legend.text = element_text(size=20),
        legend.title = element_text(size=20),
        axis.line = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = .5, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain"),
        axis.text.y = element_text(color = "grey20", size = 25, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain"),
        strip.text.y = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain")
  )

@@

# probably the closest to what others have done:
richness_mod_2 <- glmmTMB(
  Richness ~ Year_factor*Habitat*Site + (1 | Site/Transect/Location) + ar1(Year_factor + 0 | Location),
  family = poisson,
  data = alpha_diversity_quad,
  control = glmmTMBControl(optCtrl = list(iter.max = 2000, eval.max = 2000))
) # overspecified but going with it for now

summary(richness_mod_2) # looks like strong autocorrelation, which makes sense to me
Anova(richness_mod_2)
plot(allEffects(richness_mod_2))
# emmeans(richness_mod_2, pairwise ~ Habitat | Year_factor) # if we want to compare habitats year to year

```
```{r turnover troubleshooting, fig.height = 30, fig.width=25}

@@

alpha_diversity_quad_sub <- alpha_diversity_quad[!alpha_diversity_quad$Location %in% problem_children, ]

alpha_diversity_quad_sub$Cover <- rowSums(data_sub[ , -which(names(data_sub) %in% meta_cols)])/100
alpha_diversity_quad_sub$Stratum <- 
  paste(alpha_diversity_quad_sub$Site, "-", alpha_diversity_quad_sub$Habitat, "-", alpha_diversity_quad_sub$Transect)
lm_synchrony_quadrat <- codyn::synchrony(df = long_data_sub,
                                   time.var = "Year",
                                   species.var = "Taxa",
                                   abundance.var = "Percent_cover",
                                   metric = "Loreau",
                                   replicate.var = "Location") # Loreau and Mazancourt synchrony
colnames(lm_synchrony_quadrat) = c("Location", "Loreau_synchrony") # rename cols for merge later

@@

(shannon_cv_plot_2 = 
  ggplot(data = diversity_stability_quadrat) +
  geom_point(aes(x = Shannon_mean, y = Cover_cv, colour = Site, shape = Habitat), size = 4, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +
  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         )

@@

```{r portfolio effects and synchrony transect level,  fig.height = 8, fig.width = 10}
(loreau_stability_plot_quad = 
  ggplot(data = diversity_stability_quadrat) +
  geom_point(aes(x = Loreau_synchrony, y = Cover_cv, colour = Site, shape = Habitat), size = 4, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +

@@

(portfolio_stability_plot_quad = 
  ggplot(data = diversity_stability_quadrat) +
  geom_point(aes(x = Portfolio_effect, y = Cover_cv, colour = Site, shape = Habitat), size = 8, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +

@@

```{r diversity portfolio effects and synchrony quadrat level,  fig.height = 8, fig.width = 10}
(loreau_diversity_plot_quad = 
  ggplot(data = diversity_stability_quadrat) +
  geom_point(aes(x = Shannon_mean, y = Loreau_synchrony, colour = Site, shape = Habitat), size = 4, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +

@@

>>>>

(richness_quad_plot <-
  ggplot(data = alpha_diversity_quad_means) +
  geom_point(aes(x = Year, y = Richness_mean, colour = Site, shape = Habitat, group = Site_habitat), size = 8) +
  geom_line(aes(x = Year, y = Richness_mean, colour = Site, group = Site_habitat), size = 1.2) +
  geom_errorbar(aes(x = Year, ymin = Richness_mean - Richness_se , ymax = Richness_mean + Richness_se, colour = Site,
                    group = Site_habitat), width = 0.5, size = 1.2) +
  scale_colour_manual(values = site_colours) +
  scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Habitat)+
  labs(title = "a. Richness quad level\n", y = "Richness\n") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        axis.line.x = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(size = 32),
        legend.key.size = unit(2, "cm"), # increase legend size
        legend.text = element_text(size = rel(2)), # increase legend text
        legend.title = element_text(size = rel(2.25)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal',
        legend.box.background = element_rect(colour = "black") # make legend horizontal
        )
    )

@@

slope_mod <- glmmTMB(Slope ~ Habitat*Site + (1|Site/Transect), data = gls_coefficients_merged)
# model convergence problem
summary(slope_mod) # pretty low variance in random effects
plot(gls_res <- simulateResiduals(slope_mod)) # there are issues, but I can't figure out what they are...
# I think this is due to 1200 obs in model
hist(residuals(slope_mod))
plot(predict(slope_mod) ~ residuals(slope_mod))
Anova(slope_mod)
# Response: Slope
#                 Chisq Df Pr(>Chisq)    
# Habitat      1397.999  3  < 2.2e-16 ***
# Site           39.095  5  2.273e-07 ***
# Habitat:Site 1738.155 15  < 2.2e-16 ***
r.squaredGLMM(slope_mod) # 0.22, 0.24
plot(allEffects(slope_mod))
# Backreef: richness increases on East and North shore, but relatively constant west side
# Fringe: richness decreasing on east and west side, but relatvely constant on north shore
# Outer 10: Richness neutral/decreasing on east side, but neutral/increasing on north and west side
# Outer 17: Richness decreasing north/east side, but neutral west side
# Worth noting trends are slight - arguable whether they are biologically meaningful

slope_emm <- emmeans(slope_mod, ~ Habitat*Site) # create emmeans object
emmeans::contrast(slope_emm, "consec", simple = "each", combine = TRUE, adjust = "bonferroni")
# Note: not sure I want to include this
emmeans(slope_mod, pairwise ~ Site | Habitat)
```

@@

ggplot(data = gls_coefficients_merged) +
  geom_histogram(aes(x = Slope), bins = 30) +
  geom_vline(xintercept = 0, size = 2, linetype = 2) +
  facet_grid(Habitat ~ Site) +
#  theme_classic() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
#        panel.background = element_blank(),
        legend.text = element_text(size=20),
        legend.title = element_text(size=20),
        axis.line = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = .5, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain"),
        axis.text.y = element_text(color = "grey20", size = 25, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain"),
        strip.text.y = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain")
  )

@@

# probably the closest to what others have done:
richness_mod_2 <- glmmTMB(
  Richness ~ Year_factor*Habitat*Site + (1 | Site/Transect/Location) + ar1(Year_factor + 0 | Location),
  family = poisson,
  data = alpha_diversity_quad,
  control = glmmTMBControl(optCtrl = list(iter.max = 2000, eval.max = 2000)), na.action = na.pass
) # overspecified but going with it for now

summary(richness_mod_2) # looks like strong autocorrelation, which makes sense to me
Anova(richness_mod_2)
plot(allEffects(richness_mod_2))
# emmeans(richness_mod_2, pairwise ~ Habitat | Year_factor) # if we want to compare habitats year to year
# model.sel(dredge(richness_mod_2))

top_rich_mod <- glmmTMB(Richness ~ Year_factor*Habitat + Year_factor*Site + (1 | Site/Transect/Location) + ar1(Year_factor + 0 | Location),   family = poisson, data = alpha_diversity_quad, control = glmmTMBControl(optCtrl = list(iter.max = 2000, eval.max = 2000)))
summary(top_rich_mod)
Anova(top_rich_mod)
```
```{r turnover troubleshooting, fig.height = 30, fig.width=25}

@@

alpha_diversity_quad$Cover <- rowSums(data[ , -which(names(data) %in% meta_cols)])/100
alpha_diversity_quad_sub <- alpha_diversity_quad[!alpha_diversity_quad$Location %in% problem_children, ]

# alpha_diversity_quad_sub$Cover <- rowSums(data_sub[ , -which(names(data_sub) %in% meta_cols)])/100
alpha_diversity_quad_sub$Stratum <- 
  paste(alpha_diversity_quad_sub$Site, "-", alpha_diversity_quad_sub$Habitat, "-", alpha_diversity_quad_sub$Transect)
lm_synchrony_quadrat <- codyn::synchrony(df = long_data_sub,
                                   time.var = "Year",
                                   species.var = "Taxa",
                                   abundance.var = "Percent_cover",
                                   metric = "Loreau",
                                   replicate.var = "Location") # Loreau and Mazancourt synchrony
colnames(lm_synchrony_quadrat) = c("Location", "Loreau_synchrony") # rename cols for merge later

@@

(shannon_cv_plot_2 = 
  ggplot(data = diversity_stability_quadrat) +
  geom_point(aes(x = Shannon_mean, y = Cover_cv, colour = Site, shape = Habitat), size = 4, stroke = 1) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Habitat) +
  theme_bw() + 
  theme(axis.title.x = element_text(size=18), 
        axis.title.y = element_text(size=18), 
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.text = element_text(size = rel(1)), # increase legend text
        legend.title = element_text(size = rel(1)),
        legend.position = 'bottom', # move legend to bottom
        legend.box = 'horizontal', # make legend horizontal
        legend.box.background = element_rect(colour = "black"))  # add box around legend
         )

@@

```{r portfolio effects and synchrony transect level,  fig.height = 8, fig.width = 10}
(loreau_stability_plot_quad = 
  ggplot(data = diversity_stability_quadrat) +
  geom_point(aes(x = Loreau_synchrony, y = Cover_cv, colour = Site, shape = Habitat), size = 4, stroke = 1) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Habitat) +

@@

(portfolio_stability_plot_quad = 
  ggplot(data = diversity_stability_quadrat) +
  geom_point(aes(x = Portfolio_effect, y = Cover_cv, colour = Site, shape = Habitat), size = 4, stroke = 2) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Site) +

@@

```{r diversity portfolio effects and synchrony quadrat level,  fig.height = 8, fig.width = 10}
(loreau_diversity_plot_quad = 
  ggplot(data = diversity_stability_quadrat) +
  geom_point(aes(x = Shannon_mean, y = Loreau_synchrony, colour = Site, shape = Habitat), size = 4, stroke = 1) + 
   scale_colour_manual(values = site_colours) +
   scale_shape_manual(values = habitat_shapes) +
  facet_wrap(~Habitat) +

@@

```{r diversity models quad level}
diversity_mod_universal <- glmmTMB(Cover_cv ~Shannon_mean*Site + Shannon_mean*Habitat + Site*Habitat + (1|Transect/Location), family = Gamma(link = "log"), data = diversity_stability_quadrat, na.action = na.pass)
diversity_dredge <- dredge(diversity_mod_universal)
model.sel(diversity_dredge) # best model is as specified
summary(diversity_mod_universal) # low variance in random effects
Anova(diversity_mod_universal) # LR Tests are kinda useless
r.squaredGLMM(diversity_mod_universal) # 0.653, 0.655

synchrony_mod_universal <- glmmTMB(Cover_cv ~Loreau_synchrony*Site + Loreau_synchrony*Habitat + Site*Habitat + (1|Transect/Location), family = Gamma(link = "log"), data = diversity_stability_quadrat, na.action = na.pass)
synchrony_dredge <- dredge(synchrony_mod_universal)
model.sel(synchrony_dredge) # best model is as specified
summary(synchrony_mod_universal)
r.squaredGLMM(synchrony_mod_universal) # 0.800, 0.802

diversity_stability_quadrat$Loreau_synchrony_trans <- sv_trans(diversity_stability_quadrat$Loreau_synchrony, diversity_stability_quadrat) # for modelling on beta
synchrony_diversity_universal <- glmmTMB(Loreau_synchrony_trans ~Shannon_mean*Site + Shannon_mean*Habitat + Site*Habitat + (1|Transect/Location), family = beta_family(), data = diversity_stability_quadrat, na.action = na.pass)
model.sel(dredge(synchrony_diversity_universal)) # best model as specified
summary(synchrony_diversity_universal)
Anova(synchrony_diversity_universal)
r.squaredGLMM(synchrony_diversity_universal) # Model's distribution-specific variance is negative. Results are not reliable.
